---
title: "Extinction risk from climate change: geography"
output: word_document
date: "Mar. 25, 2024"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache.lazy = FALSE) 
```

# Load libraries and data
```{r load libraries and data, warning=FALSE}

rm(list = ls())
 root.dir = "C:/Users/mcu08001/Documents/1New Research/CC MetaRisk2/Analysis"

 #load libraries
library(coda); library(ggplot2); library(rstan); library(bayesplot); library(shinystan); library(loo); library(rstanarm); library(dplyr); library(ggpubr)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = FALSE)

#load data
dataP<-read.table("MetaRisk2 aggthres 5.txt",header=T); attach(dataP)

#Bayesian stan model proportional and weighted
#betareg requires no 0s or 1s
koffset = 0.001 #the k that gives the best posterior predictive check
percent2 <- adj.percent
percent2[adj.percent == 0] = koffset;
percent2[adj.percent == 1] = 1 - koffset;
dataP$percent2 <- percent2;
```

## Intercept-only model
First need to calculate the intercept-only model (mean) to compare against categorical models.

```{r intercept model}
data.use<-dataP

N = length(data.use$percent2)
n.Study <- length(unique(data.use$Study)) #number of studies
Studyint<-as.integer(unclass(factor(data.use$Study)))
phi = data.use$Total.N

#create model matrix for coefficients
stan.data<-list(N = N, percent = data.use$percent2, phi = phi, S = n.Study, Study = Studyint)
params.to.monitor=c("mu","y_rep","study_u", "psi","eta","log_lik")

init.fn<- function (chain_id) {
  list(mu = -2.5)
}
load("2all_interc.rds") # includes all data *dataP*
# mod=stan(file="MetaRisk2 RSTAN int only 1.stan",data=stan.data,pars=params.to.monitor,
#          chains = 3, warmup=5000, cores=3,iter=8000, save_warmup = FALSE,
#          init = init.fn, control=list(adapt_delta = 0.9, max_treedepth = 15))

params.to.monitor2=c("mu")#
sumx = summary(mod,probs=c(.025,0.975), digits=4, pars=params.to.monitor2) 
sumx$summary

#checks
traceplot(mod,pars=params.to.monitor2,inc_warmup=FALSE)

pp_check(
  stan.data$percent,
  rstan::extract(mod, par = 'y_rep')$y_rep[1:100, ], 
  fun = 'dens_overlay'
) 

#calculate loo
# log_lik_1 <- extract_log_lik(mod, merge_chains = FALSE)
# r_eff <- relative_eff(exp(log_lik_1), cores = 6)
# loo.mod <- loo(log_lik_1, r_eff = r_eff, cores = 6)
loo.mod # 
```

## Test for effect of Region 
Here I tested if extinction predictions change based on region - represented by continent and including marine habitats. 

```{r geography analysis}
data.use<-dataP

N = length(data.use$percent2)
n.Study <- length(unique(data.use$Study)) #number of studies
Studyint<-as.integer(unclass(factor(data.use$Study)))
phi = data.use$Total.N

#create model matrix for coefficients
betamat <- (model.matrix(~Region,data=data.use, contrasts.arg=list(Region="contr.sum"))) #create model matrix 1-Africa, 2-Asia, 3-New Zealand/Australia, 4-Europe, 5-marine, 6-N.America, 7-S.America, 8-world
stan.data<-list(N = N, percent = data.use$percent2, betamat = betamat, phi = phi, S = n.Study, P = ncol(betamat), Study = Studyint)

params.to.monitor=c("beta","y_rep","stu","sigma_stu", "eta","log_lik")

init.fn<- function (chain_id) {
  list(beta = c(-2.5, rep(0,ncol(betamat)-1)))
}

# mod=stan(file="MetaRisk2 RSTAN beta mat.stan",data=stan.data,pars=params.to.monitor,
#           chains = 3, warmup=18000, cores=3,iter=21000,
#          init = init.fn, save_warmup = FALSE, control=list(adapt_delta = 0.9, max_treedepth = 20))#

load("2geog5.rds") #previous version had divergences, this one run 18000 - 21000
params.to.monitor2=c("beta")#
sumx = summary(mod,probs=c(.025,0.975), digits=4, pars=params.to.monitor2) 
sumx$summary

#checks
traceplot(mod,pars=params.to.monitor2,inc_warmup=FALSE)

pp_check(
  stan.data$percent,
  rstan::extract(mod, par = 'y_rep')$y_rep[1:100, ], 
  fun = 'dens_overlay'
) 

#calculate loo
# log_lik_1 <- extract_log_lik(mod, merge_chains = FALSE)
# r_eff <- relative_eff(exp(log_lik_1), cores = 6)
# loo.mod <- loo(log_lik_1, r_eff = r_eff, cores = 6)
loo.mod # 
```

```{r LOO table}
#create data frame of looics from two models
load("2geog5.rds")
loo.mod2=loo.mod # rename loo.mod so can load n
mod2 = mod
#load("beta interc only.rds")
load("2all_interc.rds") #use full data (dataP) intercept only model for comparison
 
table.data<-data.frame(
  Model = c("Intercept-only model","Model including geography"),
  LOOic = c(loo.mod$estimates[3],loo.mod2$estimates[3]),
  SE = c(loo.mod$estimates[6],loo.mod2$estimates[6])
)
loo.diff = loo.mod2$estimates[3] - loo.mod$estimates[3]
print(loo.diff)
knitr::kable(table.data, caption = "Table x: Comparisons of LOOic between baseline and Model with geography", format = "markdown")
```

#LOOIC results


```{r, calculate and table}
load("2geog5.rds")

modx = mod
cats <- c("Africa", "Asia", "New Zealand/Australia", "Europe", "Oceans", "North America", "South America", "Global")
#Calculate estimates; note original is 1 in matrix
posterior=as.data.frame(modx); 

#absolute risks
gm<-posterior[["beta[1]"]]
beta.1<-posterior[["beta[1]"]]+posterior[["beta[2]"]]
beta.2<-posterior[["beta[1]"]]+posterior[["beta[3]"]]
beta.3<-posterior[["beta[1]"]]+posterior[["beta[4]"]]
beta.4<-posterior[["beta[1]"]]+posterior[["beta[5]"]]
beta.5<-posterior[["beta[1]"]]+posterior[["beta[6]"]]
beta.6<-posterior[["beta[1]"]]+posterior[["beta[7]"]]
beta.7<-posterior[["beta[1]"]]+posterior[["beta[8]"]]
beta.8<-8*gm -(beta.1+beta.2+beta.3+beta.4+beta.5+beta.6+beta.7) #last level is missing in sum contrasts so needs to be calculated from rest

#use common median threat
load("2all_interc.rds")
posterior2=as.data.frame(mod)
grand.mean = posterior2[["mu"]]
grand.mean.inv =invlogit(grand.mean)
grand.mean.pred = quantile(grand.mean.inv,probs = c(0.025, 0.5, 0.975))
grand.mean.pred

beta.cat<-cbind(beta.1,beta.2,beta.3,beta.4,beta.5,beta.6,beta.7,beta.8)
#calculate median and credible intervals
pred.cat = invlogit(apply(beta.cat, 2, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.cat.df <- data.frame(x = cats,
                          mean = pred.cat[2,],
                          low = pred.cat[1,],
                          hi = pred.cat[3,])

#arrange in increasing order#
pred.cat.df$cats = with (pred.cat.df, reorder(cats, mean)) 
pred.cat.df <- pred.cat.df[pred.cat.df$cats != "Oceans",] #remove Oceans, which is covered by habitat grouping now

#relative risks
rbeta.1<-posterior[["beta[2]"]]
rbeta.2<-posterior[["beta[3]"]]
rbeta.3<-posterior[["beta[4]"]]
rbeta.4<-posterior[["beta[5]"]]
rbeta.5<-posterior[["beta[6]"]]
rbeta.6<-posterior[["beta[7]"]]
rbeta.7<-posterior[["beta[8]"]]
rbeta.8<-beta.8 - grand.mean

rbeta.cat = invlogit(beta.cat)-invlogit(grand.mean)
rpred.cat = (apply(rbeta.cat, 2, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
rpred.cat.df <- data.frame(x = cats,
                          mean = rpred.cat[2,],
                          low = rpred.cat[1,],
                          hi = rpred.cat[3,])
#Differences from zero
glob.mean.over <- rep("n",nrow(rpred.cat.df))
glob.mean.over[rpred.cat.df$mean > 0 & rpred.cat.df$low > 0] = "y"
glob.mean.over[rpred.cat.df$mean < 0 & rpred.cat.df$hi < 0] = "y"
rpred.cat.df$glob.mean.over <- glob.mean.over

#arrange in increasing order#
rpred.cat.df$cats = with (rpred.cat.df, reorder(cats, mean)) 
rpred.cat.df <- rpred.cat.df[rpred.cat.df$cats != "Oceans",]

#calculate global mean
g.mean = grand.mean.pred[2]

pred.cat.df.r <- pred.cat.df[,1:4]
pred.cat.df.r[,2:4] <- round((pred.cat.df.r[,2:4]), digits = 4)
knitr::kable(pred.cat.df.r, caption = "Table x: Extinction risk by continent and difference from mean, with 95% credible intervals", format = "markdown", font = 5)
```

```{r Fig 1: Geography, fig.cap = "Fig. 1. Predicted extinction risk based on continent."}
#Figures
Fig1a<-ggplot(data = pred.cat.df)+
  geom_vline(xintercept=g.mean) +
  geom_errorbar(aes(y = cats, xmin = low, xmax = hi), width = 0) +
  geom_point(stat = "identity", aes(y = cats, x = mean), color = "#416788", size = 3, shape = 15) +
  xlab("Pre-industrial rise \n in temperature (C)") + xlim(c(0,.26)) +
  theme_classic()+
  theme(axis.title.y = element_blank(),axis.title=element_text(size=14),axis.text = element_text(size=12))+
  guides(size=F)
Fig1a

Fig1b<-ggplot(data = rpred.cat.df)+
  geom_vline(xintercept=0) +
  geom_errorbar(aes(y = cats, xmin = low, xmax = hi), width = 0) +
  geom_point(stat = "identity", aes(y = cats, x = mean, color = glob.mean.over), size = 3, shape = 15) +
  scale_color_manual(values=c('grey','#E98a15'))+
  xlab("Percent difference \n from global mean") + xlim(c(-.1,.2)) +
  theme_classic()+
  theme(axis.title.y = element_blank(),axis.ticks.y = element_blank(),axis.line.y = element_blank(),axis.text.y = element_blank(),
        axis.title=element_text(size=14),axis.text = element_text(size=12),legend.position = "none")+
  guides(size=F) 
Fig1b

ggarrange(Fig1a, NULL, Fig1b, ncol=3, widths = c(4,.4, 2))

ggsave("Metarisk2 geog.png",width=6,height=5,unit="in",dpi="print")

(sum(dataP$Region == "Namerica") + sum(dataP$Region == "Europe"))/nrow(dataP)

```



```{r sample sizes}
#catalog sample sizes
N.1.st<-length(unique(dataP$Study[dataP$Region == "Asia"]))
N.1.mod<-length((dataP$Study[dataP$Region == "Asia"]))

N.3.st<-length(unique(dataP$Study[dataP$Region == "Samerica"]))
N.3.mod<-length((dataP$Study[dataP$Region == "Samerica"]))

N.4.st<-length(unique(dataP$Study[dataP$Region == "World"]))
N.4.mod<-length((dataP$Study[dataP$Region == "World"]))

N.5.st<-length(unique(dataP$Study[dataP$Region == "Europe"]))
N.5.mod<-length((dataP$Study[dataP$Region == "Europe"]))

N.6.st<-length(unique(dataP$Study[dataP$Region == "Namerica"]))
N.6.mod<-length((dataP$Study[dataP$Region == "Namerica"]))

N.7.st<-length(unique(dataP$Study[dataP$Region == "Africa"]))
N.7.mod<-length((dataP$Study[dataP$Region == "Africa"]))

N.8.st<-length(unique(dataP$Study[dataP$Region == "AustraliaNewZealand"]))
N.8.mod<-length((dataP$Study[dataP$Region == "AustraliaNewZealand"]))

table.data<-data.frame(
  Factor = c("Asia","South America","Global","Europe","North America","Africa","Australia/NewZealand"),
  Studies = c(N.1.st,N.3.st,N.4.st,N.5.st,N.6.st,N.7.st,N.8.st),
  Models = c(N.1.mod,N.3.mod,N.4.mod,N.5.mod,N.6.mod,N.7.mod,N.8.mod)
)
knitr::kable(table.data, caption = "Table 2: Number of studies and models for each factor", format = "markdown")
```
## Conclusion
Geographic areas vary in risk, with the highest risks in South America, Australia, and New Zealand. The least risks are in Asia, but this is likely due to a bias toward studies in temperate Asia and focus on trees.




