---
title: "Factor-weighted estimate of extinction risk from climate change"
output: word_document
date: "March 25, 2024"
---
## Load libraries and data

```{r, libraries and data,  message = F}
knitr::opts_chunk$set(echo = TRUE, cache.lazy = FALSE, cache = TRUE) 
rm(list = ls())
 root.dir = "C:/Users/mcu08001/Documents/1New Research/CC MetaRisk2/Analysis"
library(MCMCglmm); library(coda); library(ggplot2); library(rstan); library(bayesplot); library(shinystan); library(loo); library(rstanarm);

dataP<-read.table("Metarisk2 aggthres 5.txt",header=T); #newest data 
dataP2<-dataP[is.finite(dataP$Pre.Ind.Rise),]; attach(dataP2) # need to eliminate NA s for pre-industrial rise or stat programs crash

#Bayesian stan model proportional and weighted
#betareg requires no 0s or 1s
koffset = 0.001 #the k that gives the best posterior predictive check
percent2 <- adj.percent
percent2[adj.percent == 0] = koffset;
percent2[adj.percent == 1] = 1 - koffset;
dataP2$percent2 <- percent2;

data.use<-dataP2

#other quantities
P.Ind<-seq(from = 0, to = 5.5,by = .1) #prediction interval and frequency
P.Ind2<-seq(from = 0, to = 5.5,by = .01) #prediction interval and frequency needed for Fig 1b
N = length(data.use$percent2)
n.Study <- length(unique(data.use$Study)) #number of studies
betamat <- (model.matrix(~Pre.Ind.Rise:Region,data=data.use)) 
```

## Load analyses

```{r load baseline}
#load baseline overall analysis
load("2pre_lowb.rds") #baseline
base<-mod;

#Calculate estimates; note original is 1 in matrix
posteriorb=as.data.frame(base);
mu.b<-posteriorb[["mu"]]
beta.b1<-posteriorb[["beta"]]
params.to.monitor2=c("mu","beta")#
sumx = summary(base,probs=c(.025,0.975), digits=4, pars=params.to.monitor2) 
sumx$summary
```
### Load continent analysis

```{r continent analysis}
# dataP<-read.table("MetaRisk2 aggthres 5.txt",header=T); 
# dataP2<-dataP[is.finite(dataP$Pre.Ind.Rise),]; attach(dataP2) # need to eliminate NA s for pre-industrial rise or stat programs crash
# 
# #betareg requires no 0s or 1s
# koffset = 0.001 #the k that gives the best posterior predictive check
# percent2 <- adj.percent
# percent2[adj.percent == 0] = koffset;
# percent2[adj.percent == 1] = 1 - koffset;
# dataP2$percent2 <- percent2;
# 
# data.use<-dataP2
# 
# N = length(data.use$percent2)
# n.Study <- length(unique(data.use$Study)) #number of studies
# Studyint<-as.integer(unclass(factor(data.use$Study)))
# phi = data.use$Total.N
# 
# #create model matrix for coefficients
# betamat <- (model.matrix(~Pre.Ind.Rise:Region,data=data.use)) 
# betamat.noint <- betamat[,2:ncol(betamat)]  #remove intercept and model separately
# stan.data<-list(N = N, percent = data.use$percent2, betamat = betamat.noint, phi = phi, S = n.Study, P = ncol(betamat.noint), Study = Studyint)
# 
# params.to.monitor=c("mu","beta","y_rep","stu","sigma_stu", "eta","log_lik")
# 
# init.fn<- function (chain_id) {
#   list(beta = c(-5, rep(0.5,ncol(betamat))))
# }

# mod=stan(file="MetaRisk2 RSTAN quad.stan",data=stan.data,pars=params.to.monitor,
#           chains = 3, warmup=18000, cores=3,iter=21000,
#          init = init.fn, save_warmup = FALSE, control=list(adapt_delta = 0.9, max_treedepth = 20))#


#load continent analysis
load("2geog_pre.rds") 
cont<-mod;
posterior=as.data.frame(cont);
cats <- c("Africa", "Asia", "New Zealand/Australia", "Europe", "North America", "South America")

params.to.monitor2=c("beta")#params.to.monitor2=c("mu","beta")#
sumx = summary(cont,probs=c(.025,0.975), digits=4, pars=params.to.monitor2) 
sumx$summary
```

```{r continent richness}
#Calculate estimates; note original is 1 in matrix
mu<-posterior[["mu"]]
beta.1<-posterior[["beta[1]"]]
beta.2<-posterior[["beta[2]"]]
beta.3<-posterior[["beta[3]"]]
beta.4<-posterior[["beta[4]"]]
#beta.5<-posterior[["beta[5]"]] #oceans not treated  here
beta.6<-posterior[["beta[6]"]]
beta.7<-posterior[["beta[7]"]] #no beta 8 which is world (mixed)

beta.cat<-cbind(beta.1,beta.2,beta.3,beta.4,beta.6,beta.7)
```

### Perform weighting

```{r Continent Weighting}
#Preliminary weighting based on trees (Gatti el. 2022, using observed https://www.pnas.org/doi/epdf/10.1073/pnas.2115329119)
tree.rich <- c(10441,7035,6680,7035,8646,27186) #modeled eurasia so split species evenly between Asia and Europe; used n species from Oceania for Australia/New Zealand
vert.rich <- c(4646 ,6597,1973,550, 4085,5620)#amphibians, mammals, reptiles from IUCN ranges
# africa	asia	australia	oceania	south_america	europe	north_america
# 1132	1603	222	73	2369	94	1180
# 1364	1926	271	113	1202	260	923
# 2150	3068	961	333	2049	196	1982
# 4646	6597	1454	519	5620	550	4085
cont.rich = tree.rich + vert.rich;
rel.rich <- cont.rich/sum(cont.rich);
rel.cont.richm = matrix(rep(rel.rich,9000), nrow = 9000, byrow = T)

cont.rich.db <- data.frame(
  Continents = cats,
  Total.Rich = cont.rich,
  Rel.Rich = rel.rich
)

knitr::kable(cont.rich.db, caption = "Table 1: Continental richness", format = "markdown", font = 8)

beta.cont = beta.cat * rel.cont.richm #weighted betas
beta.cont.w = rowSums(beta.cont) #sum to get weighted coefficient (works because all relative to 1)

diff.betas <- quantile(invlogit(beta.cont.w) - invlogit(beta.b1), probs = c(0.025, 0.5, 0.975))
diff.mus <- quantile(invlogit(mu) - invlogit(mu.b), probs = c(0.025, 0.5, 0.975))
cat("Difference in betas =",diff.betas)
print("")
cat("Difference in intercepts =", diff.mus)
```
## Results
The model weighted by continental richness was characterized by a larger slope and smaller intercept.

```{r, estimates}
#Calculate estimates
pred.reg = sapply(1:length(mu.b), FUN = function(x) {mu.b[x] + beta.b1[x]*P.Ind}) #mu[x]
pred.reg.quant = invlogit(apply(pred.reg, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df <- data.frame(x = P.Ind,
                          mean_line = pred.reg.quant[2,],
                          low_line = pred.reg.quant[1,],
                          hi_line = pred.reg.quant[3,])

pred.regw = sapply(1:length(beta.1), FUN = function(x) {mu[x] + beta.cont.w[x]*P.Ind}) #mu[x]
pred.reg.quantw = invlogit(apply(pred.regw, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df$mean_linew = pred.reg.quantw[2,]
pred.reg.df$low_linew = pred.reg.quantw[1,]
pred.reg.df$hi_linew = pred.reg.quantw[3,]

#----finer data ------#
pred.reg2 = sapply(1:length(mu.b), FUN = function(x) {mu.b[x] + beta.b1[x]*P.Ind2}) #mu[x]
pred.reg.quant2 = invlogit(apply(pred.reg2, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df2 <- data.frame(x = P.Ind2,
                          mean_line = pred.reg.quant2[2,],
                          low_line = pred.reg.quant2[1,],
                          hi_line = pred.reg.quant2[3,])

 
pred.regw2 = sapply(1:length(beta.1), FUN = function(x) {mu[x] + beta.cont.w[x]*P.Ind2}) #mu[x]
pred.reg.quantw2 = invlogit(apply(pred.regw2, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df2$mean_linew = pred.reg.quantw2[2,]
pred.reg.df2$low_linew = pred.reg.quantw2[1,]
pred.reg.df2$hi_linew = pred.reg.quantw2[3,]
```

```{r Fig 1, continent diversity, warning = F}

Fig1<-ggplot(data = pred.reg.df)+
  geom_ribbon(data = pred.reg.df, aes(x=P.Ind,ymin=low_line,ymax=hi_line),alpha=.2,fill="darkred") +
  geom_line(data = pred.reg.df, aes(x=P.Ind,y=mean_line),size=3,color="darkred") +
    geom_ribbon(data = pred.reg.df, aes(x=P.Ind,ymin=low_linew,ymax=hi_linew),alpha=.2,fill="#416788") +
  geom_line(data = pred.reg.df, aes(x=P.Ind,y=mean_linew),size=3,color="#416788") +
  xlab("Pre-industrial rise in temperature (C)") + ylab("Predicted extinction risk") +
  theme_classic() + scale_x_continuous(breaks = seq(0,5,1), limits = c(0,5.5)) + scale_y_continuous(breaks = seq(0,.45,0.05), limits = c(0,.45)) +
  theme(axis.title=element_text(size=18),title=element_text(size=20),axis.text = element_text(size=16))+
  guides(size=F)
Fig1

#ggsave("Fig contwtd preind.png",width=8,height=6,unit="in",dpi="print")
```
### Despite differences in intercepts and slopes, the two estimates overlap acros the range of values and 95% credible intervals incorporate the alternative estimate. The bias-corrected estimate has larger credible intervals, likely reflecting the higher uncertainties in more diverse regions such as South America.

## Now look at taxonomic richness

```{r taxonomic richness}
# #create model matrix for coefficients
# betamat <- (model.matrix(~Pre.Ind.Rise:Taxa,data=data.use)) 
# betamat.noint <- betamat[,2:ncol(betamat)]  #remove intercept and model separately
# stan.data<-list(N = N, percent = data.use$percent2, betamat = betamat.noint, phi = phi, S = n.Study, P = ncol(betamat.noint), Study = Studyint)
# 
# params.to.monitor=c("mu","beta","y_rep","stu","sigma_stu", "eta","log_lik")
# 
# init.fn<- function (chain_id) {
#   list(beta = c(-5, rep(0.5,ncol(betamat))))
# }
# 
# mod=stan(file="MetaRisk2 RSTAN quad.stan",data=stan.data,pars=params.to.monitor,
#           chains = 3, warmup=18000, cores=3,iter=21000,
#          init = init.fn, save_warmup = FALSE, control=list(adapt_delta = 0.9, max_treedepth = 20))#
# save(mod, loo.mod, file="2taxa_pre.rds")
#load taxa analysis
load("2taxa_pre2.rds") 
tax<-mod;
tposterior=as.data.frame(tax);
tcats <- c("Amphibians", "Birds", "Fish", "Insects", "Invertebrates", "Mammals", "Plants", "Reptiles")

params.to.monitor2=c("mu","beta")#
sumx = summary(tax,probs=c(.025,0.975), digits=4, pars=params.to.monitor2) 
sumx$summary

#Calculate estimates; note original is 1 in matrix
tmu<-tposterior[["mu"]]
tbeta.1<-tposterior[["beta[1]"]] #amphibs
tbeta.2<-tposterior[["beta[2]"]] #Birds
tbeta.3<-tposterior[["beta[3]"]] #Fish
tbeta.4<-tposterior[["beta[4]"]] #Insects
tbeta.5<-tposterior[["beta[5]"]] #Inverts
tbeta.6<-tposterior[["beta[6]"]] #Mammals
tbeta.7<-tposterior[["beta[7]"]] #Mixed
tbeta.8<-tposterior[["beta[8]"]] #Plants
tbeta.9<-tposterior[["beta[9]"]] #Reptiles

tbeta.cat<-cbind(tbeta.1,tbeta.2,tbeta.3,tbeta.4,tbeta.5,tbeta.6,tbeta.8,tbeta.9) # missing tbeta.7 = mixed taxa, not informative
```

```{r taxonomic Weighting}
#         "Amphibians", "Birds", "Fish", "Insects", "Invertebrates", "Mammals", "Plants", "Reptiles"
taxa.rich <- c(6515,9990,31269,1000000,359365,5487,297857,8734)
rel.taxa.rich <- taxa.rich/sum(taxa.rich);
rel.taxa.richm = matrix(rep(rel.taxa.rich,9000), nrow = 9000, byrow = T)

taxa.rich.db <- data.frame(
  taxa = tcats,
  Total.Rich = taxa.rich,
  Rel.Rich = rel.taxa.rich
)

knitr::kable(taxa.rich.db, caption = "Table 2: Taxa richness", format = "markdown", font = 8)

beta.tax = tbeta.cat * rel.taxa.richm  #weighted betas
beta.tax.w = rowSums(beta.tax) #sum to get weighted coefficient (works because all relative to 1)

diff.tbetas <- quantile(invlogit(beta.tax.w) - invlogit(beta.b1), probs = c(0.025, 0.5, 0.975))
diff.tmus <- quantile(invlogit(tmu) - invlogit(mu.b), probs = c(0.025, 0.5, 0.975))
cat("Taxa: Difference in betas =",diff.tbetas)
print(",")
cat("Taxa: Difference in intercepts =", diff.tmus)
```

## Results
Again, the weighted estimate has higher beta but lower intercept.

```{r, taxa estimates}
tpred.regw = sapply(1:length(tbeta.1), FUN = function(x) {tmu[x] + beta.tax.w[x]*P.Ind}) #mu[x]
tpred.reg.quantw = invlogit(apply(tpred.regw, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df$mean_linet = tpred.reg.quantw[2,]
pred.reg.df$low_linet = tpred.reg.quantw[1,]
pred.reg.df$hi_linet = tpred.reg.quantw[3,]

tpred.regw2 = sapply(1:length(tbeta.1), FUN = function(x) {tmu[x] + beta.tax.w[x]*P.Ind2}) #mu[x]
tpred.reg.quantw2 = invlogit(apply(tpred.regw2, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df2$mean_linet = tpred.reg.quantw2[2,]
pred.reg.df2$low_linet = tpred.reg.quantw2[1,]
pred.reg.df2$hi_linet = tpred.reg.quantw2[3,]
```

```{r Fig 2, taxa diversity}
Fig2<-ggplot(data = pred.reg.df)+
  geom_ribbon(data = pred.reg.df, aes(x=P.Ind,ymin=low_line,ymax=hi_line),alpha=.2,fill="darkred") +
  geom_line(data = pred.reg.df, aes(x=P.Ind,y=mean_line),size=3,color="darkred") +
    geom_ribbon(data = pred.reg.df, aes(x=P.Ind,ymin=low_linet,ymax=hi_linet),alpha=.2,fill="#416788") +
  geom_line(data = pred.reg.df, aes(x=P.Ind,y=mean_linet),size=3,color="#416788") +
  xlab("Pre-industrial rise in temperature (C)") + ylab("Predicted extinction risk") +
  theme_classic() + scale_x_continuous(breaks = seq(0,5,1), limits = c(0,5.5)) + scale_y_continuous(breaks = seq(0,.45,0.05), limits = c(0,.45)) +
  theme(axis.title=element_text(size=18),title=element_text(size=20),axis.text = element_text(size=16))+
  guides(size=F)
Fig2

#ggsave("Fig taxawtd preind.png",width=8,height=6,unit="in",dpi="print")
```

## Combine the two estimates into ensemble

```{r, combine estimates}
#Combine both estimates
tg.predw <- cbind(pred.regw,tpred.regw)
tg.reg.quantw = invlogit(apply(tg.predw, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df$mean_linetg = tg.reg.quantw[2,]
pred.reg.df$low_linetg = tg.reg.quantw[1,]
pred.reg.df$hi_linetg = tg.reg.quantw[3,]

tg.predw2 <- cbind(pred.regw2,tpred.regw2)
tg.reg.quantw2 = invlogit(apply(tg.predw2, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df2$mean_linetg = tg.reg.quantw2[2,]
pred.reg.df2$low_linetg = tg.reg.quantw2[1,]
pred.reg.df2$hi_linetg = tg.reg.quantw2[3,]

#save(pred.reg.df,pred.reg.df2,file = "preind Weighted estimate.rds")
```

```{r Fig 3, both weighted analyses}
Fig3<-ggplot(data = pred.reg.df)+
  geom_ribbon(data = pred.reg.df, aes(x=P.Ind,ymin=low_line,ymax=hi_line),alpha=.2,fill="darkred") +
  geom_line(data = pred.reg.df, aes(x=P.Ind,y=mean_line),size=3,color="darkred") +
    geom_ribbon(data = pred.reg.df, aes(x=P.Ind,ymin=low_linetg,ymax=hi_linetg),alpha=.2,fill="#416788") +
  geom_line(data = pred.reg.df, aes(x=P.Ind,y=mean_linetg),size=3,color="#416788") +
  xlab("Pre-industrial rise in temperature (C)") + ylab("Predicted extinction risk") +
  theme_classic() + scale_x_continuous(breaks = seq(0,5,1), limits = c(0,5.5)) + scale_y_continuous(breaks = seq(0,.45,0.05), limits = c(0,.45)) +
  theme(axis.title=element_text(size=18),title=element_text(size=20),axis.text = element_text(size=16))+
  guides(size=F)
Fig3

#ggsave("Fig bothwtd preind.png",width=8,height=6,unit="in",dpi="print")
```