---
title: "Extinction risk from climate change: latitude"
output: word_document
date: "March 15, 2024"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache.lazy = FALSE) 
```

# Load libraries and data

```{r load libraries and data, message = FALSE,warning=FALSE}
rm(list = ls())
 root.dir = "C:/Users/mcu08001/Documents/1New Research/CC MetaRisk2/Analysis"

#load libraries
library(coda); library(ggplot2); library(rstan); library(bayesplot); library(shinystan); library(loo); library(rstanarm); library(dplyr); library(ggpubr)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = FALSE)

#load data
dataP<-read.table("MetaRisk2 aggthres 5.txt",header=T); attach(dataP)

#other quantities
#betareg requires no 0s or 1s
koffset = 0.001 #the k that gives the best posterior predictive check
percent2 <- adj.percent
percent2[adj.percent == 0] = koffset;
percent2[adj.percent == 1] = 1 - koffset;
dataP$percent2 <- percent2;
data.use<-dataP

# Other quantities
N = length(data.use$percent2)
n.Study <- length(unique(data.use$Study)) #number of studies
Studyint<-as.integer(unclass(factor(data.use$Study)))
phi = data.use$Total.N

length(unique(dataP$Study))
```

## Test for effect of latitudinal band

Here I tested if extinction predictions change based on latitudinal band, represented by Antarctic, southern middle latitudes, tropics, northern middle latitudes, and Arctic regions. I predict that tropical regions will be characterized by the highest extinction risk.

```{r latidudinal bands analysis}
betamat <- (model.matrix(~Antarctic + S.Middle + Tropics + N.Middle + Arctic, data=data.use, contrasts.arg=list(Antarctic = "contr.sum", S.Middle = "contr.sum", Tropics="contr.sum", N.Middle = "contr.sum", Arctic = "contr.sum"))) #relative to global mean 

stan.data<-list(N = N, percent = data.use$percent2, betamat = betamat, phi = phi, S = n.Study, P = ncol(betamat), Study = Studyint)
params.to.monitor=c("beta","y_rep","stu","sigma_stu", "eta","log_lik")

init.beta=rep(0,ncol(betamat)-1)
init.fn<- function (chain_id) {
  list(beta = c(-2.5,init.beta))
}

# mod=stan(file="MetaRisk2 RSTAN beta mat.stan",data=stan.data,pars=params.to.monitor,
#           chains = 3, warmup=7000, cores=3,iter=10000,
#          init = init.fn, save_warmup = FALSE, control=list(adapt_delta = 0.9, max_treedepth = 15))#

load("2all_lat.rds")
params.to.monitor2=c("beta")#
sumx = summary(mod,probs=c(.025,0.975), digits=4, pars=params.to.monitor2) 
sumx$summary

#checks
traceplot(mod,pars=params.to.monitor2,inc_warmup=FALSE)

pp_check(
  stan.data$percent,
  rstan::extract(mod, par = 'y_rep')$y_rep[1:100, ], 
  fun = 'dens_overlay'
) 

#calculate loo
# log_lik_1 <- extract_log_lik(mod, merge_chains = FALSE)
# r_eff <- relative_eff(exp(log_lik_1), cores = 6)
# loo.mod <- loo(log_lik_1, r_eff = r_eff, cores = 6)
loo.mod # 
```

```{r LOO table 1}
#create data frame of looics from two models
load("2all_lat.rds")
loo.mod2=loo.mod # rename loo.mod so can load n

mod2 = mod
load("2all_interc.rds")

table.data<-data.frame(
  Model = c("Intercept-only model","Model including latitude"),
  LOOic = c(loo.mod$estimates[3],loo.mod2$estimates[3]),
  SE = c(loo.mod$estimates[6],loo.mod2$estimates[6])
)
knitr::kable(table.data, caption = "Table 1: Comparisons of LOOic between baseline and Model with latitudinal bands", format = "markdown")
Looic.diff = loo.mod2$estimates[3] - loo.mod$estimates[3]
cat("Difference in LOOic =", Looic.diff)
```

## Results

Latitudinal variation in extinction risk is not supported by model comparison.

```{r compile individual results}
# dataP<-read.table("MetaRisk2 aggthres 5.txt",header=T); 
# attach(dataP) # need to eliminate NA s for pre-industrial rise or stat programs crash
# 
# #betareg requires no 0s or 1s
# koffset = 0.001 #the k that gives the best posterior predictive check
# percent2 <- adj.percent
# percent2[adj.percent == 0] = koffset;
# percent2[adj.percent == 1] = 1 - koffset;
# dataP$percent2 <- percent2;
# 
# data.use<-dataP
# 
# #use one
# betamat <- model.matrix(~Antarctic, data = data.use, contrasts.arg = list(Antarctic = "contr.sum")) # did models with all and individually. However, because factors are not independent, 
# betamat <- model.matrix(~S.Middle, data = data.use, contrasts.arg = list(S.Middle = "contr.sum")) # did models with all and individually. However, because factors are not independent, 
# betamat <- model.matrix(~Tropics, data = data.use, contrasts.arg = list(Tropics = "contr.sum")) # did models with all and individually. However, because factors are not independent, 
# betamat <- model.matrix(~N.Middle, data = data.use, contrasts.arg = list(N.Middle = "contr.sum")) # did models with all and individually. However, because factors are not independen
# betamat <- model.matrix(~Arctic, data = data.use, contrasts.arg = list(Arctic = "contr.sum")) # did models with all and individually. However, because factors are not independent, 
# 
# N = length(data.use$percent2)
# n.Study <- length(unique(data.use$Study)) #number of studies
# Studyint<-as.integer(unclass(factor(data.use$Study)))
# phi = data.use$Total.N
# 
# #create model matrix for coefficients
# stan.data<-list(N = N, percent = data.use$percent2, betamat = betamat, phi = phi, S = n.Study, P = ncol(betamat), Study = Studyint)
# 
# params.to.monitor=c("beta","y_rep","stu","sigma_stu", "eta","log_lik")
# 
# init.beta=rep(0,ncol(betamat)-1)
# init.fn<- function (chain_id) {
#   list(beta = c(-2.5,init.beta))
# }
# 
# mod=stan(file="MetaRisk2 RSTAN beta mat.stan",data=stan.data,pars=params.to.monitor,
#           chains = 3, warmup=5000, cores=3,iter=8000,
#          init = init.fn, save_warmup = FALSE, control=list(adapt_delta = 0.9, max_treedepth = 15))#

n.total <- nrow(data.use)  #total N

#use common median threat
load("2all_interc.rds")
posterior2=as.data.frame(mod)
grand.mean = posterior2[["mu"]]
grand.mean.inv =invlogit(grand.mean)
r.mean.1 = grand.mean.inv
grand.mean.pred = quantile(grand.mean.inv,probs = c(0.025, 0.5, 0.975))
grand.mean.pred

load("2antarct.rds")
mod.Ant <- mod
posterior <- as.data.frame(mod.Ant)
p.Ant <- sum(Antarctic == "Y")/n.total #prop of yes
p.other <- sum(Antarctic == "N")/n.total #prop of nos
beta.Y = invlogit(posterior[["beta[1]"]] - posterior[["beta[2]"]])
beta.N = invlogit(posterior[["beta[1]"]] + posterior[["beta[2]"]])
#r.mean.1 = (p.other * beta.N + p.Ant * beta.Y)
beta.vs.r.mean.1 = (beta.Y) - (r.mean.1)

param.vals.1 <- data.frame(
"r.mean.1" = r.mean.1,
"beta.N" = beta.N,
"beta.Y" = beta.Y,
"beta.vs.r.mean.1" = beta.vs.r.mean.1
)
pred.1 = (apply(param.vals.1, 2, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))

region.vals <- data.frame(
  region = "Antarctic",
  beta.Y.lo = pred.1[1,3],
  beta.Y = pred.1[2,3],
  beta.Y.hi = pred.1[3,3],
  beta.vs.mean.lo = pred.1[1,4],
  beta.vs.mean = pred.1[2,4],
  beta.vs.mean.hi = pred.1[3,4]
)
#*****************************************************************************************
load("2s_middle.rds")
#load("beta s.Middle.rds")
mod.SM <- mod
reg <- S.Middle
posterior <- as.data.frame(mod.SM)

p.Y <- sum(reg == "Y")/n.total #prop of yes
p.other <- sum(reg == "N")/n.total #prop of nos
beta.Y = invlogit(posterior[["beta[1]"]] - posterior[["beta[2]"]])
beta.N = invlogit(posterior[["beta[1]"]] + posterior[["beta[2]"]])
#r.mean.1 = (p.other * beta.N + p.Y * beta.Y)
beta.vs.r.mean.1 = beta.Y - r.mean.1

param.vals.1 <- data.frame(
"r.mean.1" = r.mean.1,
"beta.N" = beta.N,
"beta.Y" = beta.Y,
"beta.vs.r.mean.1" = beta.vs.r.mean.1
)
pred.1 = (apply(param.vals.1, 2, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
region.vals[2,] <- cbind("S.Middle", t(pred.1[,3]), t(pred.1[,4]))

#*********************************************************************************************
load("2tropics.rds")
#load("beta tropics.rds")
mod.trop <- mod 
reg <- Tropics
posterior <- as.data.frame(mod.trop)

p.Y <- sum(reg == "Y")/n.total #prop of yes
p.other <- sum(reg == "N")/n.total #prop of nos
beta.Y = invlogit(posterior[["beta[1]"]] - posterior[["beta[2]"]])
beta.N = invlogit(posterior[["beta[1]"]] + posterior[["beta[2]"]])
#r.mean.1 = (p.other * beta.N + p.Y * beta.Y)
beta.vs.r.mean.1 = beta.Y - r.mean.1

param.vals.1 <- data.frame(
"r.mean.1" = r.mean.1,
"beta.N" = beta.N,
"beta.Y" = beta.Y,
"beta.vs.r.mean.1" = beta.vs.r.mean.1
)
pred.1 = (apply(param.vals.1, 2, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
region.vals[3,] <- cbind("Tropics", t(pred.1[,3]), t(pred.1[,4]))
#*********************************************************************************************

load("2n_middle.rds")
mod.NM <- mod 
reg <- N.Middle
posterior <- as.data.frame(mod.NM)

p.Y <- sum(reg == "Y")/n.total #prop of yes
p.other <- sum(reg == "N")/n.total #prop of nos
beta.Y = invlogit(posterior[["beta[1]"]] - posterior[["beta[2]"]])
beta.N = invlogit(posterior[["beta[1]"]] + posterior[["beta[2]"]])
#r.mean.1 = (p.other * beta.N + p.Y * beta.Y)
beta.vs.r.mean.1 = beta.Y - r.mean.1

param.vals.1 <- data.frame(
"r.mean.1" = r.mean.1,
"beta.N" = beta.N,
"beta.Y" = beta.Y,
"beta.vs.r.mean.1" = beta.vs.r.mean.1
)
pred.1 = (apply(param.vals.1, 2, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
region.vals[4,] <- cbind("N.Middle", t(pred.1[,3]), t(pred.1[,4]))
#*********************************************************************************************

load("2arctic.rds")
mod.Arc <- mod 
reg <- Arctic
posterior <- as.data.frame(mod.Arc)

p.Y <- sum(reg == "Y")/n.total #prop of yes
p.other <- sum(reg == "N")/n.total #prop of nos
beta.Y = invlogit(posterior[["beta[1]"]] - posterior[["beta[2]"]])
beta.N = invlogit(posterior[["beta[1]"]] + posterior[["beta[2]"]])
#r.mean.1 = (p.other * beta.N + p.Y * beta.Y)
beta.vs.r.mean.1 = beta.Y - r.mean.1

param.vals.1 <- data.frame(
"r.mean.1" = r.mean.1,
"beta.N" = beta.N,
"beta.Y" = beta.Y,
"beta.vs.r.mean.1" = beta.vs.r.mean.1
)
pred.1 = (apply(param.vals.1, 2, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
region.vals[5,] <- cbind("Arctic", t(pred.1[,3]), t(pred.1[,4]))
#*********************************************************************************************
save(region.vals,file ="Latitudinal bands analysis results 2.rds")
knitr::kable(region.vals, caption = "Table 2: Extinction risk by regiona and difference from mean, with 95% credible intervals", format = "markdown", digits = 3)
```

Northern middle latitudes and Arctic appear to be characterized by lower predicted extinction risks than other regions.

```{r Fig 1: Geography, fig.cap = "Fig. 1. Predicted extinction risk based on latitudinal bands."}
#load("Latitudinal bands analysis results 2.rds")
#region.vals$region[1] = "Antarctic" #fix name
region.vals$beta.Y.lo <- as.numeric(region.vals$beta.Y.lo) # data is character and needs to be numeric
region.vals$beta.Y <- as.numeric(region.vals$beta.Y)
region.vals$beta.Y.hi <- as.numeric(region.vals$beta.Y.hi)
region.vals$beta.vs.mean.lo <- as.numeric(region.vals$beta.vs.mean.lo)
region.vals$beta.vs.mean <- as.numeric(region.vals$beta.vs.mean)
region.vals$beta.vs.mean.hi<- as.numeric(region.vals$beta.vs.mean.hi)
region.vals$region <- factor(region.vals$region, levels = c("Antarctic", "S.Middle", "Tropics", "N.Middle","Arctic"))

#Figures
Fig1a<-ggplot(data = region.vals)+
  geom_vline(xintercept=grand.mean.pred[2]) +
  geom_errorbar(aes(y = region, xmin = beta.Y.lo, xmax = beta.Y.hi), width = 0) +
  geom_point(stat = "identity", aes(y = region, x = beta.Y), color = "#416788", size = 3, shape = 15) +
  xlab("Pre-industrial rise \n in temperature (C)") + xlim(c(0,.2)) +
  theme_classic()+ scale_y_discrete(labels = c("Antarctic", "S. Middle", "Tropics", "N. Middle", "Arctic")) +
  theme(axis.title.y = element_blank(),axis.title=element_text(size=14),axis.text = element_text(size=12))+
  guides(size=F)
Fig1a

#Differences from zero
glob.mean.over <- rep("n",nrow(region.vals))
glob.mean.over[region.vals$beta.vs.mean > 0 & region.vals$beta.vs.mean.lo > 0] = "y"
glob.mean.over[region.vals$beta.vs.mean < 0 & region.vals$beta.vs.mean.hi < 0] = "y"
region.vals$glob.mean.over <- glob.mean.over

Fig1b<-ggplot(data = region.vals)+
  geom_vline(xintercept=0) +
  geom_errorbar(aes(y = region, xmin = beta.vs.mean.lo, xmax = beta.vs.mean.hi), width = 0) +
  geom_point(stat = "identity", aes(y = region, x = beta.vs.mean, color = glob.mean.over), size = 3, shape = 15) +
  scale_color_manual(values=c('grey','#5B507A'))+
  xlab("Percent difference \n from global mean") + xlim(c(-.1,.1)) +
  theme_classic()+ 
  theme(axis.title.y = element_blank(),axis.ticks.y = element_blank(),axis.line.y = element_blank(),axis.text.y = element_blank(),
        axis.title=element_text(size=14),axis.text = element_text(size=12),legend.position = "none")+
  guides(size=F) 
Fig1b

ggarrange(Fig1a, NULL, Fig1b, ncol=3, widths = c(4,.4, 2))

#ggsave("Metarisk2 lat bands.png",width=6,height=5,unit="in",dpi="print")
```

```{r sample sizes}
#catalog sample sizes
N.1.st<-length(unique(dataP$Study[dataP$Antarctic == "Y"]))
N.1.mod<-length((dataP$Study[dataP$Antarctic == "Y"]))

N.2.st<-length(unique(dataP$Study[dataP$S.Middle == "Y"]))
N.2.mod<-length((dataP$Study[dataP$S.Middle == "Y"]))

N.3.st<-length(unique(dataP$Study[dataP$Tropics == "Y"]))
N.3.mod<-length((dataP$Study[dataP$Tropics == "Y"]))

N.4.st<-length(unique(dataP$Study[dataP$N.Middle == "Y"]))
N.4.mod<-length((dataP$Study[dataP$N.Middle == "Y"]))

N.5.st<-length(unique(dataP$Study[dataP$Arctic == "Y"]))
N.5.mod<-length((dataP$Study[dataP$Arctic == "Y"]))

table.data<-data.frame(
  Factor = c("Antarctic", "S. Middle", "Tropics", "N. Middle", "Arctic"),
  Studies = c(N.1.st,N.2.st,N.3.st,N.4.st,N.5.st),
  Models = c(N.1.mod,N.2.mod,N.3.mod,N.4.mod,N.5.mod)
)
knitr::kable(table.data, caption = "Table 3: Number of studies and models for each factor", format = "markdown")
```

## Conclusion

Latitudinal bands vary in risk, with lower risk in the Northern middle latitudes and Arctic.
