---
title: "Extinction risk from climate change: ecosystem"
output: word_document
date: "Mar. 25th, 2024"
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache.lazy = FALSE, cache = TRUE) 
```
# Load libraries and data
```{r load libraries and data, include = FALSE, warning=FALSE}
rm(list = ls())
 root.dir = "C:/Users/mcu08001/Documents/1New Research/CC MetaRisk2/Analysis"
#
 #load libraries
library(coda); library(ggplot2); library(rstan); library(bayesplot); library(shinystan); library(loo); library(rstanarm); library(dplyr); library(ggpubr)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = FALSE)
#
#load data
dataP<-read.table("MetaRisk2 aggthres 5.txt",header=T); attach(dataP)
data.use<-dataP; attach(data.use)
#
#other quantities
#betareg requires no 0s or 1s
koffset = 0.001 #the k that gives the best posterior predictive check
percent2 <- adj.percent
percent2[adj.percent == 0] = koffset;
percent2[adj.percent == 1] = 1 - koffset;
dataP$percent2 <- percent2;

data.use <- dataP
attach(data.use)
N = length(data.use$percent2)
n.Study <- length(unique(data.use$Study)) #number of studies
Studyint<-as.integer(unclass(factor(data.use$Study)))
phi = data.use$Total.N

load("2all_interc.rds") #intercept only model
mod.int <- mod
loo.int <- loo.mod

#use common definition for global median for graph, or else it varies a bit
posterior2=as.data.frame(mod.int)
grand.mean = posterior2[["mu"]]
grand.mean.pred <- invlogit(quantile(grand.mean, probs = c(0.025, 0.5, 0.975)))
grand.mean.pred
```

## Test for effect of habitat
Here I tested if extinction predictions change based on habitat, including terrestrial, marine, freshwater, islands, and mountains. I predicted that extinction risks would increase on islands and mountains, and possibly freshwater also. 

```{r all habitat analysis}
#create model matrix for coefficients
data.use$habs <- ifelse(data.use$Region == "Marine","Marine",ifelse(data.use$Fresh == "Y", "Fresh","Terrestrial"))

betamat <- data.frame(
  Intercept = rep(1,N),
  Mtn = ifelse(Mtn == "Y",1,0),
  Island = ifelse(Island == "Y",1,0),
  Terrestrial = ifelse(data.use$habs == "Terrestrial",1,0),
  Marine = ifelse(data.use$habs == "Marine",1,0),
  Freshwater = ifelse(data.use$habs == "Fresh",1,0))
#
stan.data<-list(N = N, percent = data.use$percent2, betamat = betamat, phi = phi, S = n.Study, P = ncol(betamat), Study = Studyint)
params.to.monitor=c("beta","y_rep","stu","sigma_stu", "eta","log_lik")
#
init.beta=rep(0,ncol(betamat)-1)
init.fn<- function (chain_id) {
  list(beta = c(-2.5,init.beta))
}
# mod=stan(file="MetaRisk2 RSTAN beta mat.stan",data=stan.data,pars=params.to.monitor,
#           chains = 3, warmup=7000, cores=3,iter=10000,
#          init = init.fn, save_warmup = FALSE, control=list(adapt_delta = 0.9, max_treedepth = 15))#
 #          
load("2allhabs.rds")
params.to.monitor2=c("beta")#
sumx = summary(mod,probs=c(.025,0.975), digits=4, pars=params.to.monitor2) 
sumx$summary
#
#checks
traceplot(mod,pars=params.to.monitor2,inc_warmup=FALSE)
#
pp_check(
  stan.data$percent,
  rstan::extract(mod, par = 'y_rep')$y_rep[1:100, ], 
  fun = 'dens_overlay'
) 
#calculate loo
# log_lik_1 <- extract_log_lik(mod, merge_chains = FALSE)
# r_eff <- relative_eff(exp(log_lik_1), cores = 6)
# loo.mod <- loo(log_lik_1, r_eff = r_eff, cores = 6)
loo.mod # 
```

```{r LOO table 1}
#create data frame of looics from two models
load("2allhabs.rds")

table.data<-data.frame(
  Model = c("Intercept-only model","Model including habitat"),
  LOOic = c(loo.int$estimates[3],loo.mod$estimates[3]),
  SE = c(loo.int$estimates[6],loo.mod$estimates[6])
)
knitr::kable(table.data, caption = "Table 1: Comparisons of LOOic between baseline and all-habitat model", format = "markdown")
Looic.diff = loo.mod$estimates[3] - loo.int$estimates[3]
cat("Difference in LOOic =", Looic.diff)
```

## Three habitats: Marine, terrestrial and freshwater

```{r 3 habitat analysis}
#Marine is also a part of region analysis, but want to explore it as a habitat and keep region based on continent.
data.use$habs <- ifelse(data.use$Region == "Marine","Marine",ifelse(data.use$Fresh == "Y", "Fresh","Terrestrial"))
betamat <- model.matrix(~data.use$habs)
#
#Need to do each separately, because habitats overlap, e.g., freshwater on islands
#betamat <- model.matrix(~Mtn) #modeled as yes, no, some
#betamat <- model.matrix(~Island) #modeled as yes, no, some
#
stan.data<-list(N = N, percent = data.use$percent2, betamat = betamat, phi = phi, S = n.Study, P = ncol(betamat), Study = Studyint)
params.to.monitor=c("beta","y_rep","stu","sigma_stu", "eta","log_lik")
#
init.beta=rep(0,ncol(betamat)-1)
init.fn<- function (chain_id) {
  list(beta = c(-2.5,init.beta))
}
 # mod=stan(file="MetaRisk2 RSTAN betamat.stan",data=stan.data,pars=params.to.monitor,
 #           chains = 3, warmup=7000, cores=7,iter=10000,
 #          init = init.fn, control=list(adapt_delta = 0.9, max_treedepth = 15))
 #          
load("2habsL2.rds")
params.to.monitor2=c("beta")#
sumx = summary(mod,probs=c(.025,0.975), digits=4, pars=params.to.monitor2) 
sumx$summary
#
#checks
traceplot(mod,pars=params.to.monitor2,inc_warmup=FALSE)
#
pp_check(
  stan.data$percent,
  rstan::extract(mod, par = 'y_rep')$y_rep[1:100, ], 
  fun = 'dens_overlay'
) 
#calculate loo
# log_lik_1 <- extract_log_lik(mod, merge_chains = FALSE)
# r_eff <- relative_eff(exp(log_lik_1), cores = 6)
# loo.mod <- loo(log_lik_1, r_eff = r_eff, cores = 6)
loo.mod # 
```


```{r LOO table 2}
#create data frame of looics from two models
load("2habsL2.rds")
#
table.data<-data.frame(
  Model = c("Intercept-only model","Model including habitat"),
  LOOic = c(loo.int$estimates[3],loo.mod$estimates[3]),
  SE = c(loo.int$estimates[6],loo.mod$estimates[6])
)
knitr::kable(table.data, caption = "Table 2: Comparisons of LOOic between baseline and model with three habitats", format = "markdown")
Looic.diff = loo.mod$estimates[3] - loo.int$estimates[3]
cat("Difference in LOOic =", Looic.diff)
```

```{r compile individual results}
#create data frame of looics from two models
data.use<-dataP; attach(data.use)
n.total <- nrow(data.use)  #total N
#
#load("beta mtn.rds")
load("2mtn.rds")
mod.mtn <- mod
posterior <- as.data.frame(mod.mtn)
p.y <- sum(data.use$Mtn == "Y")/n.total #prop of yes
p.s <- sum(data.use$Mtn == "S")/n.total #prop of some
p.n <- sum(data.use$Mtn == "N")/n.total #prop of nos
beta.Y = invlogit(posterior[["beta[1]"]] + posterior[["beta[3]"]])
beta.S = invlogit(posterior[["beta[1]"]] + posterior[["beta[2]"]])
beta.N = invlogit(posterior[["beta[1]"]])
#r.mean.1 = p.n * beta.N + p.s * beta.S + p.y * beta.Y
r.mean.1 = invlogit(grand.mean)
beta.vs.r.mean.n = (beta.N) - (r.mean.1)
beta.vs.r.mean.s = (beta.S) - (r.mean.1)
beta.vs.r.mean.y = (beta.Y) - (r.mean.1)
#
param.vals.1 <- data.frame(
"r.mean.1" = r.mean.1,
"beta.N" = beta.N,
"beta.S" = beta.S,
"beta.Y" = beta.Y,
"beta.vs.r.mean.n" = beta.vs.r.mean.n,
"beta.vs.r.mean.s" = beta.vs.r.mean.s,
"beta.vs.r.mean.y" = beta.vs.r.mean.y
)
pred.1 = (apply(param.vals.1, 2, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
#
hab.vals <- data.frame(beta.N.lo = pred.1[1,2],
  beta.N = pred.1[2,2],
  beta.N.hi = pred.1[3,2],
  beta.S.lo = pred.1[1,3],
  beta.S = pred.1[2,3],
  beta.S.hi = pred.1[3,3], 
  beta.Y.lo = pred.1[1,4],
  beta.Y = pred.1[2,4],
  beta.Y.hi = pred.1[3,4],   
  beta.N.mean.lo = pred.1[1,5],
  beta.N.mean = pred.1[2,5],
  beta.N.mean.hi = pred.1[3,5],
  beta.S.mean.lo = pred.1[1,6],
  beta.S.mean = pred.1[2,6],
  beta.S.mean.hi = pred.1[3,6],  
  beta.Y.mean.lo = pred.1[1,7],
  beta.Y.mean = pred.1[2,7],
  beta.Y.mean.hi = pred.1[3,7]
)
#*****************************************************************************************
load("3islandAUS.rds")
mod.isl <- mod
params.to.monitor2=c("beta")#
sumx = summary(mod.isl,probs=c(.025,0.975), digits=4, pars=params.to.monitor2) 
sumx$summary
invlogit(c(-2.34,-2.34-.6,-2.34+.17))

posterior <- as.data.frame(mod.isl)
p.y <- sum(Island == "Y")/n.total #prop of yes
p.s <- sum(Island == "S")/n.total #prop of some
p.n <- sum(Island == "N")/n.total #prop of nos
beta.Y = invlogit(posterior[["beta[1]"]] + posterior[["beta[3]"]])
beta.S = invlogit(posterior[["beta[1]"]] + posterior[["beta[2]"]])
beta.N = invlogit(posterior[["beta[1]"]])
#r.mean.1 = p.n * beta.N + p.s * beta.S + p.y * beta.Y
beta.vs.r.mean.n = (beta.N) - (r.mean.1)
beta.vs.r.mean.s = (beta.S) - (r.mean.1)
beta.vs.r.mean.y = (beta.Y) - (r.mean.1)
#
param.vals.1 <- data.frame(
"r.mean.1" = r.mean.1,
"beta.N" = beta.N,
"beta.S" = beta.S,
"beta.Y" = beta.Y,
"beta.vs.r.mean.n" = beta.vs.r.mean.n,
"beta.vs.r.mean.s" = beta.vs.r.mean.s,
"beta.vs.r.mean.y" = beta.vs.r.mean.y
)
pred.1 = (apply(param.vals.1, 2, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
#
hab.vals[2,] <- cbind(beta.N.lo = pred.1[1,2],
  beta.N = pred.1[2,2],
  beta.N.hi = pred.1[3,2],
  beta.S.lo = pred.1[1,3],
  beta.S = pred.1[2,3],
  beta.S.hi = pred.1[3,3], 
  beta.Y.lo = pred.1[1,4],
  beta.Y = pred.1[2,4],
  beta.Y.hi = pred.1[3,4],   
  beta.N.mean.lo = pred.1[1,5],
  beta.N.mean = pred.1[2,5],
  beta.N.mean.hi = pred.1[3,5],
  beta.S.mean.lo = pred.1[1,6],
  beta.S.mean = pred.1[2,6],
  beta.S.mean.hi = pred.1[3,6],  
  beta.Y.mean.lo = pred.1[1,7],
  beta.Y.mean = pred.1[2,7],
  beta.Y.mean.hi = pred.1[3,7]
)
#*********************************************************************************************
load("2habsL2.rds")
mod.habs <- mod
posterior <- as.data.frame(mod.habs)

#marine
p.m <- sum(Region == "Marine")/n.total #prop of yes
p.f <- sum(Fresh == "Y")/n.total #prop of yes
p.t <- (1 - p.m - p.f) #prop of yes

beta.m = invlogit(posterior[["beta[1]"]] + posterior[["beta[2]"]])
beta.f = invlogit(posterior[["beta[1]"]])
beta.t = invlogit(posterior[["beta[1]"]] + posterior[["beta[3]"]])

#r.mean.1 = p.m * beta.m + p.f * beta.f + p.t * beta.t
beta.vs.r.mean.m = (beta.m) - (r.mean.1)
beta.vs.r.mean.f = (beta.f) - (r.mean.1)
beta.vs.r.mean.t = (beta.t) - (r.mean.1)

param.vals.1 <- data.frame(
"r.mean.1" = r.mean.1,
"beta.m" = beta.m,
"beta.f" = beta.f,
"beta.t" = beta.t,
"beta.vs.r.mean.m" = beta.vs.r.mean.m,
"beta.vs.r.mean.f" = beta.vs.r.mean.f,
"beta.vs.r.mean.t" = beta.vs.r.mean.t
)
pred.1 = (apply(param.vals.1, 2, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
nas<-c(NA,NA,NA)

hab.vals[3:5,] <- cbind(
  beta.N.lo = nas,
  beta.N = nas,
  beta.N.hi = nas,
  beta.S.lo = nas,
  beta.S = nas,
  beta.S.hi = nas, 
  beta.Y.lo = pred.1[1,2:4],
  beta.Y = pred.1[2,2:4],
  beta.Y.hi = pred.1[3,2:4],   
  beta.N.mean.lo = nas,
  beta.N.mean = nas,
  beta.N.mean.hi = nas,
  beta.S.mean.lo = nas,
  beta.S.mean = nas,
  beta.S.mean.hi = nas,  
  beta.Y.mean.lo = pred.1[1,5:7],
  beta.Y.mean = pred.1[2,5:7],
  beta.Y.mean.hi = pred.1[3,5:7]
)
#*********************************************************************************************
hab.vals$habitat = c("Mountains","Islands","Marine","Freshwater","Terrestrial")
  
#show values
hab.vals.r <- hab.vals
hab.vals.r[,1:18] <- round((hab.vals.r[,1:18]), digits = 4)
opts <- options(knitr.kable.NA = "")
knitr::kable(hab.vals.r, caption = "Table 1: Extinction risk by taxa, with 95% credible intervals", format = "markdown", font = 4, longtable = T)

```

Terrestrial habitats have lower predicted extinction risks and mountain habitats have higher predicted extinction risks than other habitats.

```{r Fig 1: Habitat, fig.cap = "Fig. 1. Predicted extinction risk based on habitat."}
#load("Habitat analysis results.rds")
hab.vals$beta.Y.lo <- as.numeric(hab.vals$beta.Y.lo) # data is character and needs to be numeric
hab.vals$beta.Y <- as.numeric(hab.vals$beta.Y)
hab.vals$beta.Y.hi <- as.numeric(hab.vals$beta.Y.hi)
hab.vals$beta.Y.mean.lo <- as.numeric(hab.vals$beta.Y.mean.lo)
hab.vals$beta.Y.mean <- as.numeric(hab.vals$beta.Y.mean)
hab.vals$beta.Y.mean.hi<- as.numeric(hab.vals$beta.Y.mean.hi)
hab.vals$habitat <- factor(hab.vals$habitat, levels = c("Terrestrial", "Freshwater", "Marine","Islands","Mountains"))

#Figures
xinter <- pred.1[2,1]
Fig1a<-ggplot(data = hab.vals)+
  geom_vline(xintercept=xinter) +
  geom_errorbar(aes(y = habitat, xmin = beta.Y.lo, xmax = beta.Y.hi), width = 0) +
  geom_point(stat = "identity", aes(y = habitat, x = beta.Y), color = "#416788", size = 3, shape = 15) +
  xlab("Pre-industrial rise \n in temperature (C)") + xlim(c(0,.2)) +
  theme_classic()+ 
  #scale_y_discrete(labels = c("Mountains","Islands","Marine","Freshwater","Terrestrial")) +
  theme(axis.title.y = element_blank(),axis.title=element_text(size=14),axis.text = element_text(size=12))+
  guides(size=F)
Fig1a

#Differences from zero
glob.mean.over <- rep("n",nrow(hab.vals))
glob.mean.over[hab.vals$beta.Y.mean > 0 & hab.vals$beta.Y.mean.lo > 0] = "y"
glob.mean.over[hab.vals$beta.Y.mean < 0 & hab.vals$beta.Y.mean.hi < 0] = "y"
hab.vals$glob.mean.over <- glob.mean.over

Fig1b <- ggplot(data = hab.vals)+
  geom_vline(xintercept=0) +
  geom_errorbar(aes(y = habitat, xmin = beta.Y.mean.lo, xmax = beta.Y.mean.hi), width = 0) +
  geom_point(stat = "identity", aes(y = habitat, x = beta.Y.mean, color = glob.mean.over), size = 3, shape = 15) +
  scale_color_manual(values=c('grey','#5B507A'))+
  xlab("Percent difference \n from global mean") + xlim(c(-.12,.12)) +
  theme_classic()+ 
  theme(axis.title.y = element_blank(),axis.ticks.y = element_blank(),axis.line.y = element_blank(),axis.text.y = element_blank(),
        axis.title=element_text(size=14),axis.text = element_text(size=12),legend.position = "none")+
  guides(size=F) 
Fig1b

ggarrange(Fig1a, NULL, Fig1b, ncol=3, widths = c(4,.4, 2))

ggsave("Metarisk2 habs.png",width=4,height=3,unit="in",dpi="print")




```


```{r Fig 2: specific results for mtns and isles, fig.cap = "Fig. 2. Predicted extinction risk based on habitats."}
threat.YN<-data.frame(
  threat = c("Mountains - N","Mountains - Y","Islands - N","Islands - Y"),
  YN = rep(c("N","Y"),2),
  beta = c(hab.vals$beta.N[1],hab.vals$beta.Y[1],hab.vals$beta.N[2],hab.vals$beta.Y[2]),
   beta.lo = c(hab.vals$beta.N.lo[1],hab.vals$beta.Y.lo[1],hab.vals$beta.N.lo[2],hab.vals$beta.Y.lo[2]),
    beta.hi = c(hab.vals$beta.N.hi[1],hab.vals$beta.Y.hi[1],hab.vals$beta.N.hi[2],hab.vals$beta.Y.hi[2])
)

#set colors
boolColors <- as.character(c("N"="#416788", "Y"="#8cb369"))
boolScale <- scale_colour_manual(name="YN", values=boolColors)

#Figures
Fig2<-ggplot(data = threat.YN)+
  geom_vline(xintercept=grand.mean.pred[2]) +
  geom_errorbar(aes(y = threat, xmin = beta.lo, xmax = beta.hi), width = 0) +
  geom_point(stat = "identity", aes(y = threat, x = beta, color = YN), size = 3, shape = 15) +
  xlab("Pre-industrial rise \n in temperature (C)") + xlim(c(0.05,.20)) +
  theme_classic() + boolScale + 
  theme(axis.title.y = element_blank(),axis.title=element_text(size=14),axis.text = element_text(size=12))+
  guides(color = guide_legend(""))
Fig2

ggsave("Metarisk2 hab YN.png",width=6,height=5,unit="in",dpi="print")
```

```{r sample sizes}
#catalog sample sizes
habs <- ifelse(data.use$Region == "Marine","Marine",ifelse(data.use$Fresh == "Y", "Fresh","Terrestrial"))
#
N.mtn.st<-length(unique(dataP$Study[dataP$Mtn == "Y"]))
N.mtn.mod<-length((dataP$Study[dataP$Mtn == "Y"]))

N.isl.st<-length(unique(dataP$Study[dataP$Island == "Y" | dataP$Region == "AustraliaNewZealand"]))
N.isl.mod<-length((dataP$Study[dataP$Island == "Y" | dataP$Region == "AustraliaNewZealand"]))

N.mar.st<-length(unique(dataP$Study[dataP$Region == "Marine"]))
N.mar.mod<-length((dataP$Study[dataP$Region == "Marine"]))

N.fresh.st<-length(unique(dataP$Study[dataP$Fresh == "Y"]))
N.fresh.mod<-length((dataP$Study[dataP$Fresh == "Y"]))

N.terr.st<-length(unique(dataP$Study[habs == "Terrestrial"]))
N.terr.mod<-length((dataP$Study[habs == "Terrestrial"]))

table.data<-data.frame(
  Factor = c("Mountain","Island","Marine","Freshwwater","Terrestrial"),
  Studies = c(N.mtn.st,N.isl.st,N.mar.st,N.fresh.st,N.terr.st),
  Models = c(N.mtn.mod,N.isl.mod,N.mar.mod,N.fresh.mod,N.terr.mod)
)
knitr::kable(table.data, caption = "Table 3: Number of studies and models for each factor", format = "markdown")

```

## Conclusion
Habitats vary in extinction risk, with lower risks for species in terrestrial habitats and higher risks for species in mountainous habitats.

## Variation explained

```{r, var explained, warnings = F}
#After Gelman 2019 R2 for Bayesian
#
load("3allhabsAUS.rds") #with Australia as island
posterior=as.data.frame(mod); 

#create model matrix for coefficients
data.use$habs <- ifelse(data.use$Region == "Marine","Marine",ifelse(data.use$Fresh == "Y", "Fresh","Terrestrial"))
betamat <- data.frame(
  Intercept = rep(1,N),
  Mtn = ifelse(data.use$Mtn == "Y",1,0),
  Island = ifelse(data.use$Island == "Y" | data.use$Region == "AustraliaNewZealand",1,0),
  Terrestrial = ifelse(data.use$habs == "Terrestrial",1,0),
  Marine = ifelse(data.use$habs == "Marine",1,0),
  Freshwater = ifelse(data.use$habs == "Fresh",1,0))

betamat <- as.matrix(betamat)
#Variables and matrices
S = 9000; #samples
K = ncol(betamat); #factors
p.mat <- as.matrix(posterior[,1:K])
y = dataP$percent2 
y.mat = t(matrix(rep(y,S), nrow = N, ncol = S))
y.mean <- mean(y)

#Calculate y.pred for fixed effects only
y.pred <- matrix(rep(NA, N*S), nrow = S, ncol = N)
theta <- y.pred
for (i in 1:N) {
      theta[,i] = invlogit(p.mat %*% betamat[i,])#rows = samples, cols = i
      y.pred[,i] = (theta[,i] * data.use$Total.N[i])/(theta[,i] * data.use$Total.N[i] + (1-theta[,i]) * data.use$Total.N[i])
}

#Calcluate residual variance
res.f = y.mat - y.pred    
RSS.f = rowSums((res.f)^2)
res.v.f = 1/(N -1) * RSS.f

#Calculate fit variance
pred.v.f = 1/(N-1) * rowSums((y.pred)^2)

#Calculate R2
R2.v.f = pred.v.f/(pred.v.f + res.v.f)
cat("fixed effects R2 = ", quantile(R2.v.f,probs = c(0.025, 0.5, 0.975),na.rm = T))

#Total model With random effects
y.pred.c <-(as.matrix(posterior[,(K+1):(N+K)])) #calculated in STAN, with all RE and weightings

#Calculate residual variance
res.c = y.mat - y.pred.c    
RSS.c = rowSums((res.c)^2)
res.v.c = 1/(N-1) * RSS.c

#Calculate fit variance

pred.v.c = 1/(N-1) * rowSums(y.pred.c^2)

#Calculate full model R2
R2.v.c = pred.v.c/(pred.v.c + res.v.c)
print("                                                   ")
cat("Overall model R2 = ", quantile(R2.v.c,probs = c(0.025, 0.5, 0.975),na.rm = T))

```


