---
title: "Extinction risk from climate change: through time"
output: word_document
date: "Feb. 27, 2024"
---
## Test for effect of time of study on results

Here I wanted to test to see if the year of study affected predictions. I split the data into original data from the 2015 analysis and new data. I also included an interaction term in case the slopes differ. 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache.lazy = FALSE) 
```

## Load libraries and data

```{r, load libraries and data, results='hide', message = F}
rm(list = ls())
 root.dir = "C:/Users/mcu08001/Documents/1New Research/CC MetaRisk2/Analysis"
# Load libraries and data
library(coda); library(ggplot2); library(rstan); library(bayesplot); library(loo); library(rstanarm);
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = FALSE)

dataP<-read.table("Metarisk2 aggthres 5.txt",header=T); #newest data with slight fix to dispersal data
dataP2<-dataP[is.finite(dataP$Pre.Ind.Rise),]; attach(dataP2) # need to eliminate NA s for pre-industrial rise or stat programs crash

#Bayesian stan model proportional and weighted
#betareg requires no 0s or 1s
koffset = 0.001 #the k that gives the best posterior predictive check
percent2 <- adj.percent
percent2[adj.percent == 0] = koffset;
percent2[adj.percent == 1] = 1 - koffset;
dataP2$percent2 <- percent2;

#create categorical factor of era ~ 90s, 00s, 10s, 20s
dataP2$era <- ifelse(Year < 2000, "1990s",
                     ifelse(Year < 2010, "2000s",
                            ifelse(Year < 2020, "2010s", "2020s")))
data.use<-dataP2

#Prepare data for models
N = length(data.use$percent2)
n.Study <- length(unique(data.use$Study)) #number of studies
Studyint<-as.integer(unclass(factor(data.use$Study)))
phi = data.use$Total.N
```

## Test for effect of decade
### Same-intercept model

```{r time analysis by decade, same intercept}
#create model matrix for coefficients
betamat<-(model.matrix(~era * Pre.Ind.Rise,data=data.use))[,2:8]   #create model matrix, exclude intercept which is modeled separately as mu (to allow for informed prior)

stan.data<-list(N = N, percent = data.use$percent2, betamat = betamat, phi = phi, S = n.Study, P = ncol(betamat), Study = Studyint)
# params.to.monitor=c("mu","beta","y_rep","stu","sigma_stu", "eta","log_lik")
# 
# init.fn<- function (chain_id) {
#   list(mu = -5, beta = c(0,.6,0))
# }

# mod=stan(file="MetaRisk2 RSTAN quad.stan",data=stan.data,pars=params.to.monitor,
#          chains = 3, warmup=5000, cores=7,iter=8000,
#         init = init.fn, control=list(adapt_delta = 0.9, max_treedepth = 15))

load("3timecat.rds") #minor fixes to years of publications
params.to.monitor2=c("mu","beta")#
sumx = summary(mod,probs=c(.025,0.975), digits=4, pars=params.to.monitor2) 
sumx$summary

#checks
traceplot(mod,pars=params.to.monitor2,inc_warmup=FALSE)
pp_check(
  stan.data$percent,
  rstan::extract(mod, par = 'y_rep')$y_rep[1:100, ], 
  fun = 'dens_overlay'
) 

#calculate loo
# log_lik_1 <- extract_log_lik(mod, merge_chains = FALSE)
# r_eff <- relative_eff(exp(log_lik_1), cores = 6)
# loo.mod <- loo(log_lik_1, r_eff = r_eff, cores = 6)
loo.mod # 
```


```{r LOO table 1}
load("3timecat.rds")
loo.mod2=loo.mod # rename loo.mod so can load base model

load("2pre_lowb.rds") #load base model
loo.mod1=loo.mod

#create data frame of looics from two models
table.data<-data.frame(
  Model = c("Base model","Model with decade"),
  LOOic = c(loo.mod1$estimates[3],loo.mod2$estimates[3]),
  SE = c(loo.mod1$estimates[6],loo.mod2$estimates[6])
)
knitr::kable(table.data, caption = "Comparisons of LOOic between baseline and Model with decade", format = "markdown")
Looic.diff = loo.mod2$estimates[3] - loo.mod1$estimates[3]
table.data
cat("Delta LoOic = ", Looic.diff)
```
```{r Fig 2 by decade common intercept, fig.cap = "Fig. Predicted extinction risk versus pre-industrial temperature rise across decades.",warning=FALSE}
#Prediction range
P.Ind<-seq(from = 0.4, to = 5.5,by = .1) #don't extrapolate rise  min(Pre.Ind.Rise)= 0.4

load("3timecat.rds")
modx = mod
load("2pre_lowb.rds") 

#Calculate estimates; note original is 1 in matrix
posterior=as.data.frame(modx);
a.all <- posterior[["mu"]]
b.90 <- posterior[["beta[1]"]]
b.00 <- posterior[["beta[2]"]] 
b.10 <- posterior[["beta[3]"]] 
b.20 <- posterior[["beta[4]"]] 

#For each decade
pred.reg = sapply(1:length(posterior$mu), FUN = function(x) {a.all[x] + b.90[x]*P.Ind})
pred.reg.quant = invlogit(apply(pred.reg, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df <- data.frame(x = P.Ind,
                          mean_line_90 = pred.reg.quant[2,],
                          low_line_90 = pred.reg.quant[1,],
                          hi_line_90= pred.reg.quant[3,])

pred.reg = sapply(1:length(posterior$mu), FUN = function(x) {a.all[x] + b.00[x]*P.Ind})
pred.reg.quant = invlogit(apply(pred.reg, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df$mean_line_00 = pred.reg.quant[2,]
pred.reg.df$low_line_00 = pred.reg.quant[1,]
pred.reg.df$hi_line_00= pred.reg.quant[3,]

pred.reg = sapply(1:length(posterior$mu), FUN = function(x) {a.all[x] + b.10[x]*P.Ind})
pred.reg.quant = invlogit(apply(pred.reg, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df$mean_line_10 = pred.reg.quant[2,]
pred.reg.df$low_line_10 = pred.reg.quant[1,]
pred.reg.df$hi_line_10= pred.reg.quant[3,]

pred.reg = sapply(1:length(posterior$mu), FUN = function(x) {a.all[x] + b.20[x]*P.Ind})
pred.reg.quant = invlogit(apply(pred.reg, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df$mean_line_20 = pred.reg.quant[2,]
pred.reg.df$low_line_20 = pred.reg.quant[1,]
pred.reg.df$hi_line_20= pred.reg.quant[3,]

Fig1<-ggplot(data = pred.reg.df)+
      geom_ribbon(aes(x=P.Ind,ymin=low_line_90,ymax=hi_line_90),alpha=.7,fill="#416788")+
  geom_line(aes(x=P.Ind,y=mean_line_90),size=1.5,color="#416788")+
      geom_ribbon(aes(x=P.Ind,ymin=low_line_00,ymax=hi_line_00),alpha=.7,fill="#5b507b")+
  geom_line(aes(x=P.Ind,y=mean_line_00),size=1.5,color="#5b507b")+
      geom_ribbon(aes(x=P.Ind,ymin=low_line_10,ymax=hi_line_10),alpha=.7,fill="#5b307b")+
  geom_line(aes(x=P.Ind,y=mean_line_10),size=1.5,color="#5b307b")+
  geom_ribbon(aes(x=P.Ind,ymin=low_line_20,ymax=hi_line_20),alpha=.5,fill="#5b000b")+
  geom_line(aes(x=P.Ind,y=mean_line_20),size=1.5,color="#5b000b")+
  xlab("Pre-industrial rise in temperature (C)") + ylab("Predicted extinction risk")+
  theme_classic()+
  theme(axis.title=element_text(size=18),title=element_text(size=20),axis.text = element_text(size=16))+
  guides(size="none")
Fig1

#ggsave("FigSx pre by time2.png",width=8,height=6,unit="in",dpi="print")
```



```{r time analysis by decade, full model}
#create model matrix for coefficients
betamat<-(model.matrix(~era * Pre.Ind.Rise,data=data.use))[,2:8]   #create model matrix, exclude intercept which is modeled separately as mu (to allow for informed prior)

stan.data<-list(N = N, percent = data.use$percent2, betamat = betamat, phi = phi, S = n.Study, P = ncol(betamat), Study = Studyint)
params.to.monitor=c("mu","beta","y_rep","stu","sigma_stu", "eta","log_lik")

init.fn<- function (chain_id) {
  list(mu = -5, beta = c(0,.6,0))
}

# mod=stan(file="MetaRisk2 RSTAN quad.stan",data=stan.data,pars=params.to.monitor,
#          chains = 3, warmup=5000, cores=7,iter=8000,
#         init = init.fn, control=list(adapt_delta = 0.9, max_treedepth = 15))

load("2timecat.rds")
params.to.monitor2=c("mu","beta")#
sumx = summary(mod,probs=c(.025,0.975), digits=4, pars=params.to.monitor2) 
sumx$summary

#checks
traceplot(mod,pars=params.to.monitor2,inc_warmup=FALSE)
pp_check(
  stan.data$percent,
  rstan::extract(mod, par = 'y_rep')$y_rep[1:100, ], 
  fun = 'dens_overlay'
) 

#calculate loo
# log_lik_1 <- extract_log_lik(mod, merge_chains = FALSE)
# r_eff <- relative_eff(exp(log_lik_1), cores = 6)
# loo.mod <- loo(log_lik_1, r_eff = r_eff, cores = 6)
loo.mod # 
```
```{r LOO table 2}
load("2timecat.rds")
loo.mod2=loo.mod # rename loo.mod so can load base model
load("2pre_lowb.rds") #load base model

#create data frame of looics from two models
table.data<-data.frame(
  Model = c("Base model","Model with decade (full)"),
  LOOic = c(loo.mod$estimates[3],loo.mod2$estimates[3]),
  SE = c(loo.mod$estimates[6],loo.mod2$estimates[6])
)
knitr::kable(table.data, caption = "Comparisons of LOOic between baseline and Full Model with decade", format = "markdown")
Looic.diff = loo.mod2$estimates[3] - loo.mod$estimates[3]
print(Looic.diff)
```
## Previous study versus current study
Interactive time model LOOic is larger than additive time model, delta LOOic = +6.0. Also, the interaction term overlaps with zero. No support for interactive model.

```{r time analysis (same intercept)}
#create model matrix for coefficients
betamat<-(model.matrix(~Time:Pre.Ind.Rise,data=data.use))[,2:3]  #create model matrix, exclude intercept which is modeled separately as mu (to allow for informed prior)

stan.data<-list(N = N, percent = data.use$percent2, betamat = betamat, phi = phi, S = n.Study, P = ncol(betamat), Study = Studyint)
params.to.monitor=c("mu","beta","y_rep","stu","sigma_stu", "eta","log_lik")

init.fn<- function (chain_id) {
  list(mu = -5, beta = c(0,.6,0))
}

# mod=stan(file="MetaRisk2 RSTAN quad.stan",data=stan.data,pars=params.to.monitor,
#          chains = 3, warmup=5000, cores=7,iter=8000,
#         init = init.fn, control=list(adapt_delta = 0.9, max_treedepth = 15))

load("2pre_time2.rds")
params.to.monitor2=c("mu","beta")#
sumx = summary(mod,probs=c(.025,0.975), digits=4, pars=params.to.monitor2) 
sumx$summary

#checks
traceplot(mod,pars=params.to.monitor2,inc_warmup=FALSE)

pp_check(
  stan.data$percent,
  rstan::extract(mod, par = 'y_rep')$y_rep[1:100, ], 
  fun = 'dens_overlay'
) 

#calculate loo
# log_lik_1 <- extract_log_lik(mod, merge_chains = FALSE)
# r_eff <- relative_eff(exp(log_lik_1), cores = 6)
# loo.mod <- loo(log_lik_1, r_eff = r_eff, cores = 6)
loo.mod # 
```


```{r LOO table 5}
load("2pre_time2.rds")
loo.mod2=loo.mod # rename loo.mod so can load base model
mod2 = mod
load("2pre_lowb.rds") #load base model

#create data frame of looics from two models
table.data<-data.frame(
  Model = c("Base model","Model including study period"),
  LOOic = c(loo.mod$estimates[3],loo.mod2$estimates[3]),
  SE = c(loo.mod$estimates[6],loo.mod2$estimates[6])
)
knitr::kable(table.data, caption = "Comparisons of LOOic between baseline and Model with study period", format = "markdown")
Looic.diff = loo.mod2$estimates[3] - loo.mod$estimates[3]
print(Looic.diff)
```

## Results - Analysis of original and new studies
I found no model support for an additive time factor because the change in LOOic is positive = 26.8, meaning that the model with time is worse than the model without time.

```{r Fig 1: Same intercept model, fig.cap = "Fig. x. Predicted extinction risk versus pre-industrial temperature rise for original (red) and new (blue) estimates."}
P.Ind<-seq(from = 0.4, to = 5.5,by = .1) #don't extrapolate rise  min(Pre.Ind.Rise)= 0.4

load("2pre_time2.rds")
modx = mod
load("2pre_lowb.rds") 

#Calculate estimates; note original is 1 in matrix
posterior=as.data.frame(modx); 
beta.new<-posterior[["beta[1]"]]
beta.old<-posterior[["beta[2]"]]

#compare new vs old slopes
beta.diff<-beta.new-beta.old;
beta.diff.comp <- (quantile(beta.diff, probs = c(0.025, 0.5, 0.975)))
beta.diff.comp

#For new data
pred.reg = sapply(1:length(posterior$mu), FUN = function(x) {posterior$mu[x] + beta.new[x]*P.Ind})
pred.reg.quant = invlogit(apply(pred.reg, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df <- data.frame(x = P.Ind,
                          mean_line_new = pred.reg.quant[2,],
                          low_line_new = pred.reg.quant[1,],
                          hi_line_new= pred.reg.quant[3,])

#for original data
pred.reg = sapply(1:length(posterior$mu), FUN = function(x) {posterior$mu[x] + beta.old[x]*P.Ind})
pred.reg.quant = invlogit(apply(pred.reg, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df$mean_line_orig = pred.reg.quant[2,]
pred.reg.df$low_line_orig = pred.reg.quant[1,]
pred.reg.df$hi_line_orig= pred.reg.quant[3,]

Fig1<-ggplot(data = pred.reg.df)+
      geom_ribbon(aes(x=P.Ind,ymin=low_line_orig,ymax=hi_line_orig),alpha=.7,fill="#5b507b")+
  geom_line(aes(x=P.Ind,y=mean_line_orig),size=1.5,color="#5b507b")+
  geom_ribbon(aes(x=P.Ind,ymin=low_line_new,ymax=hi_line_new),alpha=.5,fill="#416788")+
  geom_line(aes(x=P.Ind,y=mean_line_new),size=1.5,color="#416788")+
  xlab("Pre-industrial rise in temperature (C)") + ylab("Predicted extinction risk")+
  theme_classic()+
  theme(axis.title=element_text(size=18),title=element_text(size=20),axis.text = element_text(size=16))+
  guides(size=F)
Fig1

#ggsave("FigSx pre by time.png",width=8,height=6,unit="in",dpi="print")
```

```{r time analysis (same slope different intercepts)}
#create model matrix for coefficients
betamat<-(model.matrix(~Time+Pre.Ind.Rise,data=data.use))[,2:3] #create model matrix, exclude intercept which is modeled separately as mu (to allow for informed prior)

stan.data<-list(N = N, percent = data.use$percent2, betamat = betamat, phi = phi, S = n.Study, P = ncol(betamat), Study = Studyint)
params.to.monitor=c("mu","beta","y_rep","stu","sigma_stu", "eta","log_lik")

init.fn<- function (chain_id) {
  list(mu = -5, beta = c(0,.6,0))
}

# mod=stan(file="MetaRisk2 RSTAN quad.stan",data=stan.data,pars=params.to.monitor,
#          chains = 3, warmup=5000, cores=7,iter=8000,
#         init = init.fn, control=list(adapt_delta = 0.9, max_treedepth = 15))

load("2pre_time3.rds")
params.to.monitor2=c("mu","beta")#
sumx = summary(mod,probs=c(.025,0.975), digits=4, pars=params.to.monitor2) 
sumx$summary

#checks
traceplot(mod,pars=params.to.monitor2,inc_warmup=FALSE)

pp_check(
  stan.data$percent,
  rstan::extract(mod, par = 'y_rep')$y_rep[1:100, ], 
  fun = 'dens_overlay'
) 

#calculate loo
# log_lik_1 <- extract_log_lik(mod, merge_chains = FALSE)
# r_eff <- relative_eff(exp(log_lik_1), cores = 6)
# loo.mod <- loo(log_lik_1, r_eff = r_eff, cores = 6)
loo.mod # 
```


```{r LOO table 6}
load("2pre_time3.rds")
loo.mod2=loo.mod # rename loo.mod so can load base model
mod2 = mod
load("2pre_lowb.rds") #load base model

#create data frame of looics from two models
table.data<-data.frame(
  Model = c("Base model","Model including study period"),
  LOOic = c(loo.mod$estimates[3],loo.mod2$estimates[3]),
  SE = c(loo.mod$estimates[6],loo.mod2$estimates[6])
)
knitr::kable(table.data, caption = "Comparisons of LOOic between baseline and Model with study period", format = "markdown")
Looic.diff = loo.mod2$estimates[3] - loo.mod$estimates[3]
print(Looic.diff)
```

## Results - Same slope model of original and new studies
I found no model support for an additive time factor because the change in LOOic is positive = 48.9 and greater than for the same-intercept model, meaning that this model with time is worse than the model without time and with the same intercept.

```{r Same slope different intercept model, fig.cap = "Fig. x. Predicted extinction risk versus pre-industrial temperature rise for original (red) and new (blue) estimates."}
P.Ind<-seq(from = 0.4, to = 5.5,by = .1) #don't extrapolate rise  min(Pre.Ind.Rise)= 0.4

load("2pre_time3.rds")
modx = mod
load("2pre_lowb.rds") 

#Calculate estimates; note original is 1 in matrix
posterior=as.data.frame(modx); 
alpha.old<-posterior[["beta[1]"]]
beta<-posterior[["beta[2]"]]

#For new data
pred.reg = sapply(1:length(posterior$mu), FUN = function(x) {posterior$mu[x] + beta[x]*P.Ind})
pred.reg.quant = invlogit(apply(pred.reg, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df <- data.frame(x = P.Ind,
                          mean_line_new = pred.reg.quant[2,],
                          low_line_new = pred.reg.quant[1,],
                          hi_line_new= pred.reg.quant[3,])

#for original data
pred.reg = sapply(1:length(posterior$mu), FUN = function(x) {posterior$mu[x] + alpha.old[x] + beta[x]*P.Ind})
pred.reg.quant = invlogit(apply(pred.reg, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df$mean_line_orig = pred.reg.quant[2,]
pred.reg.df$low_line_orig = pred.reg.quant[1,]
pred.reg.df$hi_line_orig= pred.reg.quant[3,]

Fig1<-ggplot(data = pred.reg.df)+
      geom_ribbon(aes(x=P.Ind,ymin=low_line_orig,ymax=hi_line_orig),alpha=.7,fill="#5b507b")+
  geom_line(aes(x=P.Ind,y=mean_line_orig),size=1.5,color="#5b507b")+
  geom_ribbon(aes(x=P.Ind,ymin=low_line_new,ymax=hi_line_new),alpha=.5,fill="#416788")+
  geom_line(aes(x=P.Ind,y=mean_line_new),size=1.5,color="#416788")+
  xlab("Pre-industrial rise in temperature (C)") + ylab("Predicted extinction risk")+
  theme_classic()+
  theme(axis.title=element_text(size=18),title=element_text(size=20),axis.text = element_text(size=16))+
  guides(size=F)
Fig1

ggsave("FigSx time same slope.png",width=8,height=6,unit="in",dpi="print")
```
## Conclusion
The model with study period as a factor and its interaction with temperature does not improve upon the base model. The LOOic is larger, indicating poorer fit, however the 95% credible intervals for the slopes do not overlap zero.

## Comparing old studies and all studies

```{r comparing all and old, fig.cap = "Fig. x. Predicted extinction risk versus pre-industrial temperature rise for original (red) and all (blue) estimates."}
P.Ind<-seq(from = 0.4, to = 5.5,by = .1) #don't extrapolate rise  min(Pre.Ind.Rise)= 0.4

load("2pre_time2.rds")
modx = mod
posterior=as.data.frame(modx); 

load("2pre_lowb.rds") 
posterior.all = as.data.frame(mod)
#Calculate estimates; note original is 1 in matrix

beta.all<-posterior.all[["beta"]]
beta.old<-posterior[["beta[2]"]]

#compare new vs old slopes
beta.diff<-beta.all-beta.old;
beta.diff.comp <- (quantile(beta.diff, probs = c(0.025, 0.5, 0.975)))
beta.diff.comp

#For all data
pred.reg = sapply(1:length(posterior.all$mu), FUN = function(x) {posterior.all$mu[x] + beta.all[x]*P.Ind})
pred.reg.quant = invlogit(apply(pred.reg, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df <- data.frame(x = P.Ind,
                          mean_line_all = pred.reg.quant[2,],
                          low_line_all = pred.reg.quant[1,],
                          hi_line_all = pred.reg.quant[3,],
                          range_line_all = pred.reg.quant[3,] - pred.reg.quant[1,])

#for original data
pred.reg = sapply(1:length(posterior$mu), FUN = function(x) {posterior$mu[x] + beta.old[x]*P.Ind})
pred.reg.quant = invlogit(apply(pred.reg, 1, quantile, probs = c(0.025, 0.5, 0.975),na.rm=TRUE))
pred.reg.df$mean_line_orig = pred.reg.quant[2,]
pred.reg.df$low_line_orig = pred.reg.quant[1,]
pred.reg.df$hi_line_orig= pred.reg.quant[3,]
pred.reg.df$range_line_orig= pred.reg.quant[3,] - pred.reg.quant[1,]

pred.reg.df$prop.lower = (pred.reg.df$range_line_all)-(pred.reg.df$range_line_orig)

Fig1<-ggplot(data = pred.reg.df)+
      geom_ribbon(aes(x=P.Ind,ymin=low_line_orig,ymax=hi_line_orig),alpha=.7,fill="#5b507b")+
  geom_line(aes(x=P.Ind,y=mean_line_orig),size=1.5,color="#5b507b")+
  geom_ribbon(aes(x=P.Ind,ymin=low_line_all,ymax=hi_line_all),alpha=.5,fill="#416788")+
  geom_line(aes(x=P.Ind,y=mean_line_all),size=1.5,color="#416788")+
  xlab("Pre-industrial rise in temperature (C)") + ylab("Predicted extinction risk")+
  theme_classic()+
  theme(axis.title=element_text(size=18),title=element_text(size=20),axis.text = element_text(size=16))+
  guides(size=F)
Fig1

manual_labels <- c("range_line" = "Label A", 
                   "Group B" = "Label B",
                   "Group C" = "Label C")

Fig2<-ggplot(data = pred.reg.df)+
  geom_hline(yintercept = 0, color = "darkgrey") +
  geom_line(aes(x=P.Ind,y=range_line_orig,color="#5b507b"),size=1.5)+
  geom_line(aes(x=P.Ind,y=range_line_all,color="#416788"),size=1.5)+
  geom_line(aes(x=P.Ind,y=prop.lower/1,color="#F68928"),size=1.5)+
  scale_y_continuous(sec.axis = sec_axis(~.*1, name = "Difference in range")) +
  
  xlab("Pre-industrial rise in temperature (C)") + ylab("Credible interval range")+
  theme_classic()+  
  scale_color_identity("Credible interval range",guide = "legend", breaks = c("#5b507b","#416788","#F68928"),
                       labels =c("Original","New","Difference")) +
  theme(axis.title=element_text(size=18),title=element_text(size=20),axis.text = element_text(size=16),legend.position = c(.25, .83), legend.title = element_text(size = 18), legend.text = element_text(size = 16))
  #guides(size=F)
Fig2

#ggsave("FigS8 pre all vs old.png",width=8,height=6,unit="in",dpi="print")
```

